Supplementary Report for the article *Functional diversity can facilitate the collapse of an undesirable ecosystem state*

### 2023-01-17 reports/supplement/supplement_0.9.1.Rmd
changes supplement

We found no difference between the diversity and the tolerance manipulation when trait values were low (Figure \@ref(fig:max-tol1)). When trait values were medium to high, however, we did find differences between the diversity and tolerance manipulation for most combinations except for diversity/tolerance in only CB and in SB-PB (Figure \@ref(fig:max-tol1)). We describe and discuss these differences in detail in the sections below. Briefly, in these scenarios less tolerant strains did play a role because of their high maximum growth rates. Specifically, absence of SB strains with high growth rate led to reduced production of sulfide and therefore to coexistence with cyanobacteria at low oxygen diffusivity. In contrast, absence of PB strains with high growth rate led to reduced consumption of sulfide and therefore precluded coexistence with cyanobacteria at low oxygen diffusivity (as observed when there was high variation in PB).

Taken together, at low amounts of trait variation the diversity effects were driven entirely by the most tolerant strain, in line with a sampling effect. At medium to high trait variation, however, both tolerant and less tolerant strains affected the results in most combinations. The reason was that in the two groups of sulfur bacteria the trade-off between tolerance and growth rate had ecosystem-level effects, i.e. effects on the sulfide concentration and therefore on the cyanobacteria.

@@

At high levels of diversity / high levels of tolerance, resilience of the oxic state is greater in the tolerance than diversity manipulation. We have shown that low tolerance sulfur strains can be implicated in the oxic to anoxic transition (section \@ref(SB2))

>>>>

We found no difference between the diversity and the tolerance manipulation when trait values were low (Figure \@ref(fig:max-tol1)). When trait values were medium to high, however, we did find differences between the diversity and tolerance manipulation for most combinations except for diversity/tolerance in only CB and in SB-PB (Figure \@ref(fig:max-tol1)). Specifically, the diversity and tolerance manipulation differed either in tipping point positions (CB-SB and CB-SB-PB) or in whether coexistence patterns were atypical (SB, PB, CB-PB, CB-SB). We describe and discuss these differences in detail in the sections below. Briefly, in these scenarios less tolerant strains did play a role because of their high maximum growth rates. For example, absence of SB strains with high growth rate led to reduced production of sulfide and therefore to coexistence with cyanobacteria at low oxygen diffusivity or to increased resilience of the oxic state. In contrast, absence of PB strains with high growth rate led to reduced consumption of sulfide and therefore precluded coexistence with cyanobacteria at low oxygen diffusivity (as observed when there was high variation in PB).

Taken together, at low amounts of trait variation the diversity effects were driven entirely by the most tolerant strain, in line with a sampling effect. At medium to high trait variation, however, both tolerant and less tolerant strains affected the results in many combinations. The reason was that in the two groups of sulfur bacteria the trade-off between tolerance and growth rate had ecosystem-level effects, i.e. effects on the sulfide concentration and therefore on the cyanobacteria.

@@

At high levels of diversity / high levels of tolerance, resilience of the oxic state is greater in the tolerance than diversity manipulation. We speculate that this is because the most tolerant SB strain has reduced ability to suppress the cyanobacteria, as we have also observed in the scenarios of tolerance in only SB and in SB-CB (described above).
